Pomace: A Grappa for Non-Volatile Memory \vspace{-.2in}

### 2013-02-26 doc/nvm2013/paper.tex

\maketitle
\begin{abstract}
We revisit Grappa, our software platform that scales in-memory irregular
applications on commodity clusters.
We contemplate options for adapting Grappa to use non-volatile memory technologies.
\end{abstract}
\vspace{-.1in}

@@

\item[Communication layer.] Modern commodity networks
support high bandwidth only for large messages, but irregular applications
frequently issue small requests.   Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million 8-byte remote references per second per node.  Aggregation is largely invisible to the application programmer.

@@

\end{description}
Grappa thus facilitates
expression of fine-grained task parallelism at the application
programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.  Programmers attend only to expressing the parallelism in their program, not to scheduling it.  While similar latency-tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an \textls*{\small API} for reader-writer lock semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray \textls*{\small XMT}~\cite{feo:xmt}, a
fully custom system.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global distributed shared heap in non-volatile memory, since for irregular applications the heap
often holds data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer. 
If an 
access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address
and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
the access is to memory stored on another node, Grappa issues a
message to that node requesting the data and context switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: If a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context switch to other work
until the request had completed.

\subsection{Challenges in using commodity \textls*{\small SSD}\hspace{0.75pt}s}
Grappa's goal is to provide good irregular application performance on
mass market hardware. Unfortunately, commodity \textls*{\small SSD}\hspace{0.75pt}s are
difficult to use in this way for three reasons:

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
less than 64 bytes. Today's \textls*{\small SSD}\hspace{0.75pt}s are optimized for reads in the
kilobyte range.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores may be idle in
Grappa if each node cannot make 100 million random reads per second of
the global heap.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Commodity \textls*{\small SSD}\hspace{0.75pt}s are usually treated
as normal disks: accesses must occur through the kernel, incurring a
high context-switch overhead. Grappa depends on user-mode access to
the network layer. This approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.

@@

\subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
A natural first step would be to add one or more mass market \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
accesses and context switch to tolerate the latency of \textls*{\small SSD}
access. Unfortunately this is unlikely to work well, for two
reasons. First, as discussed earlier, the \textls*{\small SSD} must accept requests at a rate sufficient to
keep the cores busy; most \textls*{\small SSD}\hspace{0.75pt}s
can accept requests at a rate two to four orders of magnitude slower than required.
Second, the \textls*{\small SSD} must be able to support many requests operating in
parallel. Little's Law suggests that if requests take 25 $\mu$s for
the \textls*{\small SSD} to service, and if the node generates requests at a rate of
100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2500 concurrent
requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.

@@

\paragraph{Scaling down the processors.}
%\noindent{\bf Scaling down the processors.}
Alternatively, rather than scaling up the flash array, processors
could be scaled down to match the flash request rate. A low-power
processor with few cores might generate only 1 million references per
second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
There are two problems with this design for our purpose. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

\paragraph{Buffering requests to increase locality.}
%\noindent{\bf Buffering requests to increase locality.}
Grappa trades latency for better throughput by aggregating network
communications for better utilization and by migrating synchronization
operations for reduced contention. The same idea could be applied to
increase locality and amortize the cost of Flash accesses. Requests
to common flash pages would be bucketed by flash page address for a
limited time. When a sufficient number of requests have arrived for a page
or when the requests time out, a single page request would be issued to the Flash
layer; its result would be used to satisfy the individual requests.

@@

For iterative applications where the entire graph is regularly
visited, an windowed approach similar to~\cite{graphchi:osdi12} may
provide more benefit. In this approach, graph data is constantly
streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests wait to execute until the
relevant fragment of the graph is in memory. Request density is not a
concern for these applications; instead, bandwidth between the \textls*{\small SSD} and \textls*{\small DRAM}
is the limiting factor.

@@

While Grappa's goal so far has been to leverage commodity hardware, it
is worth exploring the potential benefit of emerging non-volatile
memory technologies like Phase-Change Memory. The advantages of \textls*{\small PCM}
are twofold: first, read latency is much closer to that of \textls*{\small DRAM}, and
second, \textls*{\small PCM} supports finer-grained accesses.

\paragraph{\textls*{\small PCM} as Main Memory.}
%\noindent{\bf \textls*{\small PCM} as Main Memory.}
Recent research has explored the use of \textls*{\small PCM} as a main memory
replacement~\cite{pcm:isca09}. Grappa would easily be able to take advantage of such a
development. When a Grappa program references low-locality data,
a prefetch request is issued for that data a few cycles before the
thread is context switched in. A \textls*{\small PCM}-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of \textls*{\small PCM}.

>>>>

\maketitle
\begin{abstract}
We revisit Grappa, our software platform that scales in-memory irregular
applications on mass market clusters.
We contemplate options for adapting Grappa to use non-volatile memory technologies.
\end{abstract}
\vspace{-.1in}

@@

\item[Communication layer.] Modern mass market networks
support high bandwidth only for large messages, but irregular applications
frequently issue small requests.   Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million 8-byte remote references per second per node.  Aggregation is largely invisible to the application programmer.

@@

\end{description}
Grappa thus facilitates
expression of fine-grained task parallelism at the application
programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize mass market networks.  Programmers attend only to expressing the parallelism in their program, not to scheduling it.  While similar latency-tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an \textls*{\small API} for reader-writer lock semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on mass market systems
for highly irregular applications approaching that of the Cray \textls*{\small XMT}~\cite{feo:xmt}, a
fully custom system.

@@

\section{Applying non-volatile memory to Grappa}
The natural point of integration for non-volatile memories in Grappa
is to store the global distributed shared heap in non-volatile memory, since for irregular applications the heap
often holds data that is infrequently modified. All accesses to
this shared heap in Grappa programs are done through an \textls*{\small API}, so this change could be made transparent to the programmer.

If an 
access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address
and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If
the access is to memory stored on another node, Grappa issues a
message to that node to request the data and context switches to other
work until the reply arrives. Non-volatile memory would be accessed in
the same way: If a request is to memory stored in non-volatile memory,
Grappa would initiate the request and context switch to other work
until the request had completed.

\subsection{Challenges in using mass market \textls*{\small SSD}\hspace{0.75pt}s}
Grappa's goal is to provide good irregular application performance on
mass market hardware. Unfortunately, mass market \textls*{\small SSD}\hspace{0.75pt}s are
difficult to use directly for three reasons:

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s assume locality.} In our benchmarks, we find the average request size to be
less than 64 bytes. Today's \textls*{\small SSD}\hspace{0.75pt}s are optimized for reads in the
multi-kilobyte range.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have a low request rate.} The fastest~\cite{fusionio} can
perform on the order of 1 million random reads per second, and many
are limited to less than 100K/s. Even the individual flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, cores may be idle in
Grappa if each node cannot perform 100 million random reads per second of
the global heap.

%\paragraph{\textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are usually treated
\noindent{\bf \textls*{\small SSD}\hspace{0.75pt}s have high overhead.} Mass market \textls*{\small SSD}\hspace{0.75pt}s are usually treated
as normal disks: accesses must occur through the kernel, incurring a
high context-switch overhead. Grappa depends on user-mode access to
the network layer to avoid this overhead; a similar approach has been applied to \textls*{\small SSD}\hspace{0.75pt}s in recent research~\cite{caulfield:2012}.

@@

\subsubsection{Tolerating Flash-based \textls*{\small SSD} latency} 
A natural first step would be to add one or more mass market \textls*{\small SSD}\hspace{0.75pt}s to each Grappa
cluster node, and use Grappa's global memory \textls*{\small API} to expose concurrent
accesses and context switch to tolerate the latency of \textls*{\small SSD}
access. Unfortunately this is unlikely to work well, for two
reasons. First, as discussed earlier, the \textls*{\small SSD} must accept requests at a rate sufficient to
keep the cores busy; current \textls*{\small SSD}\hspace{0.75pt}s
can accept requests at a rate two to four orders of magnitude slower than required.
Second, the \textls*{\small SSD} must be able to support many requests operating in
parallel. Little's Law suggests that if requests take 25 $\mu$s for
the \textls*{\small SSD} to service, and if the node generates requests at a rate of
100 million per second, the \textls*{\small SSD}\hspace{0.75pt}s must support 2500 concurrent
requests. Again, existing \textls*{\small SSD}\hspace{0.75pt}s support orders of magnitude less request concurrency.

@@

\paragraph{Scaling down the processors.}
%\noindent{\bf Scaling down the processors.}
Alternatively, rather than scaling up the flash array, processors
could be scaled down to match the flash request rate. A low-power
processor with few cores might generate only 1 million references per
%second. This could be satisfied with only 25\,x\,25$\mu$s Flash chips.
second. This could be satisfied with only $25\times 25\mu$s Flash chips. 
%Similar have demonstrated energy-efficient operation in recent research~\cite{gordon:2009}.
There are two problems with this design for our purpose. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip count requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

\paragraph{Buffering requests to increase locality.}
%\noindent{\bf Buffering requests to increase locality.}
Grappa trades latency for better throughput by aggregating network
communications for better utilization and by migrating synchronization
operations for reduced contention. The same idea could be applied to
increase locality and amortize the cost of Flash accesses. Requests
to nearby flash locations would be bucketed by flash page address for a
limited time. When a sufficient number of requests have arrived for a page
or when the requests time out, a single page request would be issued to the Flash
layer; its result would be used to satisfy the individual requests.

@@

For iterative applications where the entire graph is regularly
visited, an windowed approach similar to~\cite{graphchi:osdi12} may
provide more benefit. In this approach, graph data is constantly
streamed between \textls*{\small SSD} and \textls*{\small DRAM}. Requests would wait to execute until the
relevant fragment of the graph is in memory. Request density is not a
concern for these applications; instead, bandwidth between the \textls*{\small SSD} and \textls*{\small DRAM}
is the limiting factor.

@@

While Grappa's goal so far has been to leverage mass market hardware, it
is worth exploring the potential benefit of emerging non-volatile
memory technologies like Phase-Change Memory. The advantages of \textls*{\small PCM}
are twofold: first, read latency is much closer to that of \textls*{\small DRAM}, and
second, \textls*{\small PCM} supports finer-grained accesses.

\paragraph{\textls*{\small PCM} as main memory.}
%\noindent{\bf \textls*{\small PCM} as main memory.}
Recent research has explored the use of \textls*{\small PCM} as a main memory
replacement~\cite{pcm:isca09}. Grappa would easily be able to take advantage of such a
development. When a Grappa program references low-locality data,
a prefetch request is issued for that data a few cycles before the
thread is context switched in. A \textls*{\small PCM}-based
Grappa would do the same, but would issue the prefetches earlier to
cover the longer latency of \textls*{\small PCM}.
